### 2016-03-14 paper/icfp16-submission/semantics.tex
add pairs. need to update proofs

\subsection{Syntax}
\label{sec:syntax}
\input{syntax.tex}
%
Figure~\ref{fig:syntax} describes the syntax of \lang, a simple lambda
calculus with integers, booleans, and binary trees.
%
As we are specifically interested in programs that \emph{do} go wrong,
we include an explicit \stuck\ term in our syntax. We write $\estuck$ to
denote terms that may be \stuck, and $e$ to denote terms that may not be
stuck.

@@

%   % \force{v}{\subst{\tsu}{\thole}}{\vsu}{\tsu} & \mbox{if}\ \thole \in dom(\tsu) \\
%   % \triple{v}{\vsu}{\extendsu{\tsu}{\thole}{\typeof{v}}} & \mbox{otherwise} \\
% \end{cases} \\
\force{\vhole{\thole}}{t}{\vsu}{\tsu} & \defeq & \hspace{-1ex}
\begin{cases}
  \triple{v}{\vsu}{\tsu'}    & \mbox{if } v = \lookupsu{\vsu}{\ehole},
                                         \tsu' = \unify{\{\thole, t, \typeof{v}\}}{\tsu}\\
  \triple{\stuck}{\vsu}{\tsu} & \mbox{if } v = \lookupsu{\vsu}{\ehole} \\
  \triple{v}{\extendsu{\vsu}{\ehole}{v}}{\tsu'} & \mbox{if}\ \tsu' = \unify{\{\thole, t\}}{\tsu}, v = \gen{t}{\tsu'}\\
\end{cases} \\
% \begin{cases}
%   \triple{\lookupsu{\vsu}{\ehole}}{\vsu}{\tsu} & \mbox{if}\ \ehole \in dom(\vsu), \hastype{\lookupsu{\vsu}{\ehole}}{t} \\
%   \triple{\stuck}{\vsu}{\tsu}                  & \mbox{if}\ \ehole \in dom(\vsu) \\
%   \triple{v}{\extendsu{\vsu}{\ehole}{v}}{\tsu}   & v = \gen{t}\\
% \end{cases} \\
 % \begin{array}{l}
    % \mathtt{if}\ i \in \vsu\ \mathtt{then}\ \triple{\vsu(i)}{\vsu}\ \mathtt{else} \\
         % \elet{v}{\gen{t}}{\triple{v}{\ehole{i} \mapsto v}}
  % \end{array} \\
\force{n}{\tint}{\vsu}{\tsu}     & \defeq & \triple{n}{\vsu}{\tsu} \\
\force{b}{\tbool}{\vsu}{\tsu}    & \defeq & \triple{b}{\vsu}{\tsu} \\
\force{\efun{x}{e}}{\tfun}{\vsu}{\tsu} & \defeq & \triple{\efun{x}{e}}{\vsu}{\tsu} \\
\force{\vleaf{t_1}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq & \triple{\vleaf{t_1}}{\vsu}{\tsu'}, \mbox{if}\ \tsu' = \unify{\{t_1, t_2\}}{\tsu} \\
\force{\vnode{t_1}{v_1}{v_2}{v_3}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq & \triple{\vnode{t_1}{v_1}{v_2}{v_3}}{\vsu}{\tsu'}, \mbox{if}\ \tsu' = \unify{\{t_1, t_2\}}{\tsu} \\
\force{v}{t}{\vsu}{\tsu} & \defeq & \triple{\stuck}{\vsu}{\tsu}
\end{array}
$$
\caption{Narrowing values}
\label{fig:narrow}
\end{figure*}
%
%While a hole may map to a value that \emph{contains} another hole, \eg a
%lambda or a tree, it may not map \emph{directly} to another hole,
As the value substitution is normalized, in the first case of \forcesym\ we
do not need to \forcesym\ the result of the substitution, the sub-hole
will be narrowed when the context demands it.

@@

\begin{figure}[ht]
$$
\begin{array}{lcll}
\gensym       & :   & t \times \tsu \rightarrow v \\
\gen{\thole}{\tsu}  & \defeq  & \gen{\subst{\tsu}{\thole}}{\tsu}, & \quad \text{if } \thole \in dom(\tsu) \\
\gen{\tint}{\tsu}   & \defeq  & n, & \quad \text{non-deterministic} \\
\gen{\tbool}{\tsu}  & \defeq  & b, & \quad \text{non-deterministic} \\
\gen{\ttree{t}}{\tsu}  & \defeq  & tr, & \quad \text{non-deterministic} \\
\gen{\tfun}{\tsu}   & \defeq & \efun{x}{\vhole{\thole}}, & \quad \text{\ehole and \thole are fresh} \\
\gen{\thole}{\tsu}  & \defeq & \vhole{\thole}, & \quad \text{\ehole is fresh} \\
\end{array}
$$
\caption{Generating values}
\label{fig:gen}
\end{figure}

@@

\begin{figure}[ht]
\[ \begin{array}{lcll}
    \typeof{n}   & \defeq & \tint & \\
    \typeof{b}   & \defeq & \tbool & \\
    \typeof{\efun{x}{e}} & \defeq & \tfun \\
    \typeof{\vleaf{t}} & \defeq & \ttree{t} \\
    \typeof{\vnode{t}{v_1}{v_2}{v_3}} & \defeq & \ttree{t} \\
    \typeof{\vhole{\thole}} & \defeq & \thole \\
    % \typeof{\eleaf} & \defeq & \ttree{\thole}, & \quad \text{\thole is fresh} \\
    % \typeof{\enode{v_1}{v_2}{v_3}} & \defeq & \ttree{\thole}, & \quad \text{\thole is fresh} \\
    % \typeof{e} & \defeq & \thole, & \quad \text{\thole is fresh} \\
  \end{array} \]
\caption{The \emph{dynamic type} of a value.}
\label{fig:typeof}
\end{figure}

@@

\begin{cor}{\textbf{[Witness Generation]}}
\label{thm:generation}
  If $$\genWitness{k}{e} = \triple{\eapp{e}{v_1 \ldots v_n}}{\vsu}{\tsu}\ldots$$
  then for all types $t_1 \ldots t_n$ there exist values $w_1 \ldots w_n$ such that
  $\steps{\eapp{e}{w_1 \ldots w_n}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}$.
\end{cor}

>>>>

\subsection{Syntax}
\label{sec:syntax}
\input{syntax.tex}
%
Figure~\ref{fig:syntax} describes the syntax of \lang, a simple lambda
calculus with integers, booleans, pairs, and binary trees.
%
As we are specifically interested in programs that \emph{do} go wrong,
we include an explicit \stuck\ term in our syntax. We write $\estuck$ to
denote terms that may be \stuck, and $e$ to denote terms that may not be
stuck.

@@

%   % \force{v}{\subst{\tsu}{\thole}}{\vsu}{\tsu} & \mbox{if}\ \thole \in dom(\tsu) \\
%   % \triple{v}{\vsu}{\extendsu{\tsu}{\thole}{\typeof{v}}} & \mbox{otherwise} \\
% \end{cases} \\
\force{\vhole{\thole}}{t}{\vsu}{\tsu} & \defeq & \hspace{-1ex}
\begin{cases}
  \triple{v}{\vsu}{\tsu'}    & \mbox{if } v = \lookupsu{\vsu}{\ehole},
                                         \tsu' = \unify{\{\thole, t, \typeof{v}\}}{\tsu}\\
  \triple{\stuck}{\vsu}{\tsu} & \mbox{if } v = \lookupsu{\vsu}{\ehole} \\
  \triple{v}{\extendsu{\vsu}{\ehole}{v}}{\tsu'} & \mbox{if}\ \tsu' = \unify{\{\thole, t\}}{\tsu}, v = \gen{t}{\tsu'}\\
\end{cases} \\
% \begin{cases}
%   \triple{\lookupsu{\vsu}{\ehole}}{\vsu}{\tsu} & \mbox{if}\ \ehole \in dom(\vsu), \hastype{\lookupsu{\vsu}{\ehole}}{t} \\
%   \triple{\stuck}{\vsu}{\tsu}                  & \mbox{if}\ \ehole \in dom(\vsu) \\
%   \triple{v}{\extendsu{\vsu}{\ehole}{v}}{\tsu}   & v = \gen{t}\\
% \end{cases} \\
 % \begin{array}{l}
    % \mathtt{if}\ i \in \vsu\ \mathtt{then}\ \triple{\vsu(i)}{\vsu}\ \mathtt{else} \\
         % \elet{v}{\gen{t}}{\triple{v}{\ehole{i} \mapsto v}}
  % \end{array} \\
\force{n}{\tint}{\vsu}{\tsu}     & \defeq & \triple{n}{\vsu}{\tsu} \\
\force{b}{\tbool}{\vsu}{\tsu}    & \defeq & \triple{b}{\vsu}{\tsu} \\
\force{\efun{x}{e}}{\tfun}{\vsu}{\tsu} & \defeq & \triple{\efun{x}{e}}{\vsu}{\tsu} \\
\force{\epair{v_1}{v_2}}{\tprod{t_1}{t_2}}{\vsu}{\tsu} & \defeq & \triple{\epair{v_1}{v_2}}{\vsu}{\tsu''}, \mbox{if}\ \tsu' = \unify{\{\typeof{v_1}, t_1\}}{\tsu}, \tsu'' = \unify{\{\typeof{v_2}, t_2\}}{\tsu'}  \\
\force{\vleaf{t_1}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq & \triple{\vleaf{t_1}}{\vsu}{\tsu'}, \mbox{if}\ \tsu' = \unify{\{t_1, t_2\}}{\tsu} \\
\force{\vnode{t_1}{v_1}{v_2}{v_3}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq & \triple{\vnode{t_1}{v_1}{v_2}{v_3}}{\vsu}{\tsu'}, \mbox{if}\ \tsu' = \unify{\{t_1, t_2\}}{\tsu} \\
\force{v}{t}{\vsu}{\tsu} & \defeq & \triple{\stuck}{\vsu}{\tsu}
\end{array}
$$
\caption{Narrowing values}
\label{fig:narrow}
\end{figure*}
%
%While a hole may map to a value that \emph{contains} another hole, \eg a
%lambda or a tree, it may not map \emph{directly} to another hole,
As the value substitution is normalized, in the first case of \forcesym\ we
do not need to \forcesym\ the result of the substitution, the sub-hole
will be narrowed when the context demands it.

@@

\begin{figure}[ht]
$$
\begin{array}{lcll}
\gensym       & :   & t \times \tsu \rightarrow v \\
\gen{\thole}{\tsu}  & \defeq  & \gen{\subst{\tsu}{\thole}}{\tsu}, &  \text{if } \thole \in dom(\tsu) \\
\gen{\tint}{\tsu}   & \defeq  & n, &  \text{non-deterministic} \\
\gen{\tbool}{\tsu}  & \defeq  & b, &  \text{non-deterministic} \\
\gen{\tprod{t_1}{t_2}}{\tsu}  & \defeq  & \epair{\gen{t_1}{\tsu}}{\gen{t_2}{\tsu}} & \\ % \begin{subarray}{l} v_1 = \gen{t_1}{\tsu}\\ v_2 = \gen{t_2}{\tsu}\end{subarray} \\
\gen{\ttree{t}}{\tsu}  & \defeq  & tr, &  \text{non-deterministic} \\
\gen{\tfun}{\tsu}   & \defeq & \efun{x}{\vhole{\thole}}, &  \text{\ehole and \thole are fresh} \\
\gen{\thole}{\tsu}  & \defeq & \vhole{\thole}, & \text{\ehole is fresh} \\
\end{array}
$$
\caption{Generating values}
\label{fig:gen}
\end{figure}

@@

\begin{figure}[ht]
\[ \begin{array}{lcll}
    \typeof{n}   & \defeq & \tint & \\
    \typeof{b}   & \defeq & \tbool & \\
    \typeof{\efun{x}{e}} & \defeq & \tfun \\
    \typeof{\epair{v_1}{v_2}} & \defeq & \tprod{\typeof{v_1}}{\typeof{v_2}} \\
    \typeof{\vleaf{t}} & \defeq & \ttree{t} \\
    \typeof{\vnode{t}{v_1}{v_2}{v_3}} & \defeq & \ttree{t} \\
    \typeof{\vhole{\thole}} & \defeq & \thole \\
    % \typeof{\eleaf} & \defeq & \ttree{\thole}, & \quad \text{\thole is fresh} \\
    % \typeof{\enode{v_1}{v_2}{v_3}} & \defeq & \ttree{\thole}, & \quad \text{\thole is fresh} \\
    % \typeof{e} & \defeq & \thole, & \quad \text{\thole is fresh} \\
  \end{array} \]
\caption{The \emph{dynamic type} of a value.}
\label{fig:typeof}
\end{figure}

@@

\begin{cor}{\textbf{[Witness Generation]}}
\label{thm:generation}
  If $$\genWitness{k}{e} = \triple{\eapp{e}{v_1 \ldots v_n}}{\vsu}{\tsu}\ldots$$
  then for all types $t_1 \ldots t_n$ there exist values $w_1 \ldots w_n$ such that
  $\steps{\eapp{e}{w_1 \ldots w_n}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}$.
\end{cor}
\begin{proof}
  For any function $f$ of multiple arguments, we can define $f'$ to be
  the uncurried version of $f$ that takes all of its arguments as a
  nested pair, and then apply Theorem~\ref{thm:soundness} to $f'$.
\end{proof}
